---
title: "Macroscale watershed science"
subtitle: "Hacking Limnology 2023"
output:
    html_document:
        theme: "lumen"
        toc: yes
        toc_float:
            collapsed: true
---

```{r setup1, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---

## MacroSheds: long-term watershed ecosystem studies, harmonized

#### Goals:
 + enable calculation of input/output solute fluxes from diverse watersheds
 + lower barriers through data harmonization and visualization
 + investigate:
   + variation in magnitude, timing, form of N exports
   + broad effects of watershed acidification
   + variation in mineral weathering
   + watershed sensitivity to climate change
 
#### Components:
 + [macrosheds.org](https://macrosheds.org)
 + [The dataset](https://portal.edirepository.org/nis/mapbrowse?scope=edi&identifier=1262)
 + [R package](https://github.com/MacroSHEDS/macrosheds)
 + [All project code](https://github.com/MacroSHEDS)
 
#### Terms:
 + site: an individual gauging station or stream sampling location and its watershed
 + domain: one or more sites under common management
 + network: one or more domains under common funding/leadership
 + flux: solute mass normalized to watershed area, over time (kg/ha/d)
 
## Related resources

 + [GEMStat](https://gemstat.org/) - global water quality database
 + [GLORICH](https://www.geo.uni-hamburg.de/en/geologie/forschung/aquatische-geochemie/glorich.html) - global river chemistry database
 + [Water Quality Portal (WQP)](https://www.waterqualitydata.us/) - USA water quality database
 + [Global River Water Quality Archive](https://essd.copernicus.org/articles/13/5483/2021/) - subset of: GEMStat + GLORICH + WQP + CESI (Canada) + Waterbase (Europe)
 + CAMELS - Catchment Attributes and MEteorology for Large-sample Studies
   + [USA](https://gdex.ucar.edu/dataset/camels.html)
   + [Australia](https://www.webofscience.com/wos/woscc/full-record/WOS:000683780500004?SID=USW2EC0D4DYFZoLYt17MtaFngvTOg)
   + [Brazil](https://www.webofscience.com/wos/woscc/full-record/WOS:000569420300002?SID=USW2EC0D4DYFZoLYt17MtaFngvTOg)
   + [Chile](https://www.webofscience.com/wos/woscc/full-record/WOS:000449995800002?SID=USW2EC0D4DYFZoLYt17MtaFngvTOg)
   + [Great Britain](https://www.webofscience.com/wos/woscc/full-record/WOS:000580860000002?SID=USW2EC0D4DYFZoLYt17MtaFngvTOg)
 + [Caravan](https://zenodo.org/record/7944025) - subset of: CAMELS + HYSETS (N. Am.) + LamaH-CE (central Europe)
+ [CAMELS-Chem](https://hess.copernicus.org/preprints/hess-2022-81/) - chemistry for CAMELS-US

```{r setup2}
library(macrosheds)
library(tidyverse)

project_root <- '~/macrosheds_workshop'
```

```{r retrieve1}
# sites <- ms_load_sites()
# sitevars <- ms_load_variables('timeseries_by_site')

ms_download_ws_attr(project_root, dataset = 'summaries')
ws_smry <- ms_load_product(project_root, prodname = 'ws_attr_summaries')
#definitions: https://portal.edirepository.org/nis/metadataviewer?packageid=edi.1262.1 -> Data Entities -> ws_attr_summaries.csv

# site_attrs <- sites %>% 
#     filter(site_type == 'stream_gauge') %>%
#     select(site_code, ) %>% 
#     left_join(ws_smry, by = c('network', 'domain', 'site_code'))
```


# ```{r retrieve2}
# 
# ms_download_core_data()
# ms_load_product()
# macrosheds::ms_load_spatial_product()
# ```

```{r pca}
pca_data <- select(ws_smry, -site_code, -domain, -network) %>% 
    select(where( ~sum(is.na(.)) / length(.) < 0.3)) %>% #drop columns with >= 30% missing values
    drop_na() %>%                                        #drop rows with any missing values
    select(where(~sd(., na.rm = TRUE) != 0)) %>%         #drop columns with no variance
    as.matrix()

# apply(pca_data, 2, sd, na.rm = T) %>% unname()
# apply(ws_smry, 2, function(x) sum(is.na(x)))

pca <- prcomp(pca_data, center = TRUE, scale. = TRUE)
biplot(pca)
```
